An efficient composition of bidirectional programs by tupling and lazy updates

### 2019-11-19 paper/0.abst.tex
update fig + tex 20191119

 Bidirectional transformations (BX) are a solution to the view update problem and widely used for synchronizing data. The semantics and correctness of bidirectional programs has been investigated intensively during the past years, but their efficiency and optimization are not yet fully understood. In this paper, as a first step, we study different evaluation methods to optimized their evaluation. We focus on the interpretive evaluation of BX compositions because we found that these compositions are an important cause of redundant computations. 
 For evaluating BX compositions efficiently, we investigate two memoization methods. The first method, minBigul\_m, uses memoization, which improves the runtime of many BX programs by keeping intermediate results for later reuse. A disadvantage is the familiar tradeoff for keeping and searching values in a table.
  When inputs become large, the overhead increases and the effectiveness decreases. To deal with large inputs, we introduce the second method, xpg, that uses tupling, lazy update and lazy evaluation as optimizations. Lazy updates delay updates in closures and enables to use them later.
 Both evaluation methods were fully implemented for MiniBiGul and experimentally evaluated. Our experiments show the following results: the original method that is used in the language is faster for right associative compositions, minBiGUL\_m is faster for not-large inputs and xpg is faster for large inputs.

>>>>

 Bidirectional transformations (BX) are a solution to the view update problem and widely used for synchronizing data. The semantics and correctness of bidirectional programs has been investigated intensively during the past years, but their efficiency and optimization are not yet fully understood. In this paper, as a first step, we study different evaluation methods to optimized their evaluation. We focus on the interpretive evaluation of BX compositions because we found that these compositions are an important cause of redundant computations. 
 For evaluating BX compositions efficiently, we investigate two memoization methods. The first method, minBiGUL$_m$, uses memoization, which improves the runtime of many BX programs by keeping intermediate results for later reuse. A disadvantage is the familiar tradeoff for keeping and searching values in a table.
  When inputs become large, the overhead increases and the effectiveness decreases. To deal with large inputs, we introduce the second method, $xpg$, that uses tupling, lazy update and lazy evaluation as optimizations. Lazy updates delay updates in closures and enables to use them later.
 Both evaluation methods were fully implemented for minBiGUL and experimentally evaluated. Our experiments show the following results: the original method that is used in the language is faster for right associative compositions, minBiGUL$_m$ is faster for not-large inputs and $xpg$ is faster for large inputs.

### 2019-11-19 paper/1.introduction.tex
update fig + tex 20191119

In software, there are strong demands for synchronizing data. In database community this is known as ``view update problem'' and researched for a long time \cite{viewupdate}.
% A Survey to View Update Problem 
As a solution for this problem, bidirectional transformation (BX) is introduced \cite{}. As an example, let us consider a small BX program of $pHead$\footnote{The actual program is shown in the next section.}. BX provides two functions: $get$ and $put$. The following is one example of $pHead$'s evaluation.

\begin{center}
  \includegraphics[width=5cm]{./temp_fig/bxexample1.eps}
\end{center}
%\end{picture}

Here, the original source \texttt{s}, \texttt{[1,2]}, is given by a user. $get$ is a projection: $get$ of $pHead$ picks the first element of the given original source as a view.
%\texttt{get pHead [1,2] = 1}.
After the view \texttt{v} is obtained, the user can modify the view.
In this case, he or she modified the view to \texttt{8} from \texttt{1}.
From this updated view, how can we obtain new updated source? This is ``view update problem.'' For this, BX provides $put$, update on the original source.
In this example, $put$ of $pHead$ constructs a new list \texttt{[8,2]} from the updated view ($v'$) and the original source ($s$).
%\texttt{put pHead [1,2] 8 = [8,2]}

For more various evaluations, we can use composition of programs.
Let us consider another BX example program, three composition of $pHead$: $pHead \circ pHead \circ pHead$.
The following is one example of its evaluation.

\begin{center}
  \includegraphics[width=10cm]{./temp_fig/bxexample2.eps}
\end{center}

In this example, the original source \texttt{s} is \texttt{[[[1,2],[3]],[4]]}.
By repetitious evaluation of $get$s of $pHead$, we can obtain the view $v$, \texttt{1}. To obtain the final updated source \texttt{s'}, we need to evaluate $put$ three times. The first put evaluation is from \texttt{i2} and \texttt{v'} and we can obtain \texttt{i2'}.

For evaluating put-direction, there are two strategies. One is ``not keeping any intermediate states and obtaining them by evaluation when they are needed.'' In this example, ``intermediate states'' are \texttt{i1} and \texttt{i2}.
If we choose this strategy, the number of $get$s will be quadratic. This is a problem, because the evaluation will be slow if the BX programs include many compositions. In this example, two $get$s are required for obtaining \texttt{i2} and one $get$ is required for obtaining \texttt{i1}. In total, three $get$s are required.

@@

As an example, let us consider the previous example again: $pHead \circ pHead \circ pHead$.

\begin{center}
  \includegraphics[width=9cm]{./temp_fig/bxexample3.eps}
\end{center}

In the figure, \texttt{c1} and \texttt{c2} are complements. The points in this figure are two. After evaluation of the first $pg$, we do not need to keep the original source \texttt{s}, because all its information is in \texttt{c1} and \texttt{i1}. 
This evaluation looks better than previous two strategies: this does not require repeated evaluation and require smaller storage than the original sources. 
Actually, the simple combined $pg$ is not effective for left associative compositions. To achieve efficient evaluation, we introduce $cpg$, lazy update version of $pg$, in Section 4 and $kpg$, lazy evaluation version of $cpg$, in Section 5. In Section 6 we combine $pg$ and $kpg$ to achieve efficiency for both associative compositions. In Section 7 we compare all approaches including the original BX approach by the evaluation time and memory allocation.

>>>>

In software, there are strong demands for synchronizing data. In database community this is known as ``view update problem'' and researched for a long time \cite{viewupdate}.
% A Survey to View Update Problem 
As a solution for this problem, bidirectional transformation (BX) is introduced \cite{}. As an example, let us consider a small BX program of $pHead$\footnote{The actual program is shown in the next section.}. BX provides two functions: $get$ and $put$. Figure \ref{fig:eval-phead} denotes the behaviors of these functions when evaluating $phead$.

\begin{figure}[!htb]
  \begin{minipage}{0.3\textwidth}
    \centering
    \includegraphics[height=3.5cm]{./fig/fig1.png}
    \caption{Evaluating $phead$}
    \label{fig:eval-phead}
  \end{minipage}\hfill
  \begin{minipage}{0.7\textwidth}
    \centering
    \includegraphics[height=3.5cm]{./fig/fig2.png}
    \caption{Evaluating $phead \circ phead \circ phead$}
    \label{fig:eval-comp-phead}
  \end{minipage}
\end{figure}

Here, the original source \texttt{s}, \texttt{[1,2]}, is given by a user. $get$ is a projection: $get$ of $pHead$ picks the first element of the given original source as a view.
%\texttt{get pHead [1,2] = 1}.
After the view \texttt{v} is obtained, the user can modify the view.
In this case, he or she modified the view to \texttt{100} from \texttt{1}.
From this updated view, how can we obtain new updated source? This is ``view update problem.'' For this, BX provides $put$, update on the original source.
In this example, $put$ of $pHead$ constructs a new list \texttt{[100,2]} from the updated view ($v'$) and the original source ($s$).
%\texttt{put pHead [1,2] 8 = [8,2]}

For more various evaluations, we can use the composition of programs. Let us consider another example, a composition of 3 $pHead$s: $pHead \circ pHead \circ pHead$. Figure \ref{fig:eval-comp-phead} illustrates the evaluation of this program where the original source \texttt{s} is \texttt{[[[1,2],[3]],[4]]} and the updated view is \textit{100}.
By repetitious evaluation of $get$s of $pHead$, we can obtain the view $v$, \texttt{1}. To obtain the final updated source \texttt{s'}, we need to evaluate $put$ three times. The first put evaluation is from \texttt{i2} and \texttt{v'} and we can obtain \texttt{i2'}.

For evaluating put-direction, there are two strategies. One is ``not keeping any intermediate states and obtaining them by evaluation when they are needed.'' In this example, ``intermediate states'' are \texttt{i1} and \texttt{i2}.
If we choose this strategy, the number of $get$s will be linear in the number of compositions. This is a problem, because the evaluation will be slow if the BX programs include many compositions. In this example, two $get$s are required for obtaining \texttt{i2} and one $get$ is required for obtaining \texttt{i1}. In total, three $get$s are required.

@@

As an example, let's see the evaluation of $pHead \circ pHead \circ pHead$ in another way as shown in figure \ref{fig:eval-comp-phead-2}

\begin{figure}[!htb]
  \centering
  \includegraphics[height=5cm]{./fig/fig3.png}
  \caption{Evaluating $phead \circ phead \circ phead$ by keeping intermediate states}
  \label{fig:eval-comp-phead-2}
\end{figure}

In figure \ref{fig:eval-comp-phead-2}, \texttt{c1} and \texttt{c2} are complements. The points in this figure are two. After evaluation of the first $pg$, we do not need to keep the original source \texttt{s}, because all its information is in \texttt{c1} and \texttt{i1}. 
This evaluation looks better than previous two strategies: this does not require repeated evaluation and require smaller storage than the original sources. 
Actually, the simple combined $pg$ is not effective for left associative compositions. To achieve efficient evaluation, we introduce $cpg$, lazy update version of $pg$, in Section 4 and $kpg$, lazy evaluation version of $cpg$, in Section 5. In Section 6 we combine $pg$ and $kpg$ to achieve efficiency for both associative compositions. In Section 7 we compare all approaches including the original BX approach by the evaluation time and memory allocation.

### 2019-11-19 paper/2.minbigul.tex
update tex 20191114
update fig + tex 20191119

To help make demonstration more direct, we provide the following alternatives representation:

    $\tab Prod \ bx_1 \ bx_2 \equiv bx_1 \times bx_2,
    \tab Compose \ bx_1 \ bx_2 \equiv bx_1 \circ bx_2$

@@

Next, let's take a look at some examples to better understand about minBiGUL. We start with quite obivious things as follows:

$\tab \putbx{Skip \ (\lambda x.(x*x))}{10}{100} = 10 \qtab \ \getbx{Skip \ (\lambda x.(x*x))}{10} = 100\\
    \tab \putbx{Skip \ (\lambda \_.())}{1}{()} = 1 \qtabs{3} \getbx{Skip \ (\lambda \_.())}{1} = ()\\
    \tab \putbx{Replace}{1}{100} = 100 \qtabs{3} \getbx{Replace}{1} = 1
$

Now, let's consider the definition of $phead$ in minBiGUL:

>>>>

To help make demonstration more direct, we provide the following alternatives representation: 
    $\quad Prod \ bx_1 \ bx_2 \equiv bx_1 \times bx_2,
    \quad Compose \ bx_1 \ bx_2 \equiv bx_1 \circ bx_2$

@@

% Next, let's take a look at some examples to better understand about minBiGUL. We start with quite obivious things as follows:

% $\tab \putbx{Skip \ (\lambda x.(x*x))}{10}{100} = 10 \qtab \ \getbx{Skip \ (\lambda x.(x*x))}{10} = 100\\
%     \tab \putbx{Skip \ (\lambda \_.())}{1}{()} = 1 \qtabs{3} \getbx{Skip \ (\lambda \_.())}{1} = ()\\
%     \tab \putbx{Replace}{1}{100} = 100 \qtabs{3} \getbx{Replace}{1} = 1
% $

We now consider the definition of $phead$ in minBiGUL:

### 2019-11-19 paper/5.experiments.tex
update experiment 20191119
update fig + tex 20191119

\input{result2.tex}

Fig \ref{fig:result} illustrates the evaluation times of $put$ in minBiGUL, $put_m$ in minBiGUL$_m$ and $xpg$ for 12 test cases. $put$ works more efficiently in the cases of right associative compositions, rather than the left ones. This makes sense because the number of \textit{get} when evaluating a right associative composition is linear by the number of compositions, while the respective number when computing a left associative composition is significantly larger. For self-composed program, \textit{breverse}, it could be impossible to change the associate precedence from left to right. Then, $put$ works less efficiently than both $put_m$ and $xpg$. In general, for left associative compositions, if the input size is not too large, $put_m$ is the fastest method. However, if the input size is large enough, $put_m$ will be slower quickly due to times for manipulating data in the table. At that time, $xpg$ proved to be the most effective method.

>>>>

\input{result.tex}

Fig \ref{fig:result} illustrates the evaluation times of $put$ in minBiGUL, $put_m$ in minBiGUL$_m$ and $xpg$ for 12 test cases. $put$ works more efficiently in the cases of right associative compositions, rather than the left ones. This makes sense because the number of \textit{get} when evaluating a right associative composition is linear in the number of compositions, while the respective number when computing a left associative composition is significantly larger. For self-composed program, \textit{breverse}, it could be impossible to change the associate precedence from left to right. Then, $put$ works less efficiently than both $put_m$ and $xpg$. In general, for left associative compositions, if the input size is not too large, $put_m$ is the fastest method. However, if the input size is large enough, $put_m$ will be slower quickly due to times for manipulating data in the table. At that time, $xpg$ proved to be the most effective method.
